---
title: Running mixture models from Ks files
author: Morgane MILIN in collaboration with Parisod's lab
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
geometry: margin=1.5cm
output: pdf_document
---
\rule{1\linewidth}{0.5pt}

# \textcolor{blue}{Documentation}
<http://exploringdatablog.blogspot.ch/2011/08/fitting-mixture-distributions-with-r.html>  
<http://cran.r-project.org/web/packages/mixtools/vignettes/mixtools.pdf>  
\rule{1\linewidth}{0.5pt}

# \textcolor{blue}{Library}
```{r, echo=F}
library(gridExtra)
library(grid)
library(mixtools)
```
\rule{1\linewidth}{0.5pt}

# \textcolor{blue}{Parameters}
**ksmin** and **ksmax:** minimum and maximum ks value kept for the analysis  
**kmin** and **kmax:** min and max number of peaks being expected. **WARNING**: analysis time increases with k  
**boots:** bootstrapping effort during search for optimal number of peaks. **WARNING**: this is time consuming. Advised value is 1000  
**epsilon:** convergence criterion; heuristics are stopped when loglik is improved by less than epsilon  
**breaks:** number of breaks on the histogram  

```{r, echo=F}
for(arg in commandArgs()){
	tmp= unlist(strsplit(arg,split="="))
	opt = tmp[1];val = tmp[2]
	if(opt=='data_file'){infile=val}
	if(opt=='ksmin'){ksmin=as.numeric(val)}
	if(opt=='ksmax'){ksmax=as.numeric(val)}
	if(opt=='kmin'){kmin=as.numeric(val)}
	if(opt=='kmax'){kmax=as.numeric(val)}
	if(opt=='boots'){boots=as.numeric(val)}
	if(opt=='epsilon'){epsilon=as.numeric(val)}
	if(opt=='breaks'){breaks=as.numeric(val)}
}
bsn <- gsub("\\..*$", "", basename(infile))
cat("infile:", {infile}, "\n")

cat("ksmin:", {ksmin}, "; ksmax: ", {ksmax}, "\n")
cat("kmin:", {kmin}, "; kmax:", {kmax}, "\n")
cat("boots:", {boots}, "\n")
cat("epsilon:", {epsilon}, "\n")
cat("breaks:", {breaks}, "\n")
```
\rule{1\linewidth}{0.5pt}

# \textcolor{blue}{Functions}
```{r, echo=F}
plot.mixEM = function (x, whichplots = 1, loglik = 1 %in% whichplots, density = 2 %in% 
    whichplots, xlab1 = "Iteration", ylab1 = "Log-Likelihood", 
    main1 = "Observed Data Log-Likelihood", col1 = 1, lwd1 = 2, 
    xlab2 = NULL, ylab2 = NULL, main2 = NULL, col2 = NULL, lwd2 = 2, 
    alpha = 0.05, marginal = FALSE, ...){
    def.par <- par(ask = (loglik + density > 1), "mar")
    mix.object <- x
    if (!inherits(mix.object, "mixEM")) 
        stop("Use only with \"mixEM\" objects!")
    if (loglik) {
        plot(mix.object$all.loglik, xlab = xlab1, ylab = ylab1, 
            main = main1, type = "l", lwd = lwd1, col = col1, 
            ...)
    }
    if (density) {
        if (mix.object$ft == "normalmixEM") {
            k <- ncol(mix.object$posterior)
            x <- sort(mix.object$x)
            a <- hist(x, plot = FALSE, breaks = breaks)
            maxy <- max(max(a$density), 0.3989 * mix.object$lambda/mix.object$sigma)
            if (is.null(xlab2)) {
                xlab2 <- "Data"
            }
            if (is.null(col2)) {
                col2 <- 2:(k + 1)
            }
            
            ### If willing to show counts instead of densities (uncomment to use)
            a = hist(x, freq = FALSE, main = main2, xlab = xlab2, 
                ylim = c(0, maxy), ylab = ylab2, axes = F, col = "light grey", ...)
                
            ratio = max(a$density) / max(a$counts)
            labs = pretty(c(0, maxy / ratio), 8)
            labs = labs[-length(labs)]
            ats = labs * ratio
            axis(side = 1, at = round(seq(0, max(x), length.out = 5), 2))
            axis(side = 2, at = ats, label = labs)
            
            ###
            
            if (length(mix.object$mu) == 1) {
                arbvar <- TRUE
                mix.object$sigma <- mix.object$scale * mix.object$sigma
                arbmean <- FALSE
            }
            if (length(mix.object$mu) == k && length(mix.object$sigma) == 
                1) {
                arbmean <- TRUE
                arbvar <- FALSE
            }
            if (length(mix.object$sigma) == k && length(mix.object$mu) == 
                k) {
                arbmean <- TRUE
                arbvar <- TRUE
            }
            for (i in 1:k) {
                lines(x, mix.object$lambda[i] * dnorm(x, mean = mix.object$mu[i * 
                  arbmean + (1 - arbmean)], sd = mix.object$sigma[i * 
                  arbvar + (1 - arbvar)]), col = col2[i], lwd = lwd2)
            }
        }
    }
    par(def.par)
}
cat("Done\n")
```
\rule{1\linewidth}{0.5pt}


# \textcolor{blue}{Import dataset}
### \textcolor{red}{Dataset:}
```{r, echo=F}
data <- read.table(infile)
summary(data)
```

### \textcolor{red}{Filtered dataset:}
```{r, echo=F}
data2 <- data.frame(No = 1:nrow(data), Node.Ks = as.numeric(data$V4)) 	# Retrieve Ks value
data2 <- data2[!is.na(data2$Node.Ks), ]						# Remove missing Ks value
data2 <- data2[data2$Node.Ks >= ksmin & data2$Node.Ks < ksmax, ] 	# focus on a range 
summary(data2)
```

### \textcolor{red}{Filtered LOG transformed dataset:}
```{r, echo=F}
data2_log <- data2
data2_log$Node.Ks <- log(data2_log$Node.Ks)
data2_log <- data2_log[data2_log$Node.Ks != -Inf, ]
summary(data2_log)
```
\rule{1\linewidth}{0.5pt}

\newpage
# \textcolor{blue}{Estimate parameters (mixture of normal distributions)}
## \textcolor{red}{No transformation}
```{r, echo=F, fig.width=7, fig.height=7}
for (i in kmin:kmax){
	cat("Number of components:", {i}, "\n")
	
	mix.object = normalmixEM(data2_log$Node.Ks, lambda = NULL, mu = NULL, sigma = NULL, k = i, epsilon = epsilon)
	centers = round(mix.object$mu, 3); stdevs = round(mix.object$sigma, 3); contribs = round(mix.object$lambda, 3); loglik = mix.object$loglik
	params = data.frame(i, exp(centers), exp(stdevs), contribs, loglik)
	grid.arrange(tableGrob(params, theme=ttheme_minimal()))

	graph.params = hist(mix.object$x, breaks = breaks, plot = F)
	plot.mixEM(mix.object, which = 2, breaks = breaks, xlab2 = "Ks values", ylab2 = "Number of duplicates")
}
```

